Modularly Composing Typed Language Fragments

### 2014-11-03 att-pldi15/att-pldi15.tex
Compressed a bit more

@@

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We conclude by discussing  related and future work in Sec. \ref{prior-work}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only in the synthetic judgement.

@@

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation. In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.

%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}. The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
%  \begin{figure*}[t!]
% \small
% \hspace{-5px}\input{example-table}
% %\\~\\
% %$\begin{array}{rcl}
% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
% %\end{array}$
% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
% \label{example}
% \end{figure*}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}

\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-EL}
\caption{Syntax of the external language (EL).}\label{syntax-EL}
\end{figure}

@@

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined. The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}\vspace{-5px}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}\vspace{-5px}
\end{figure*}

@@

\paragraph{Examples}
Our first example of a user-defined tycon is $\tcvar{rstr}$. It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

@@

\subsection{Static Type Case Analysis}
Types in @$\lambda$ can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context \cite{conf/ppdp/LohH06}. Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. \vspace{-5px}
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}
The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (such types always take the default branch).

The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-5px}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

@@

\inferrule[k-ity-trans]{
    \sofkX{\st}{\kty}
}{
    \sofkX{\sqity{\srep{\st}}}{\kity}
}
\end{mathpar}
The unquote form is eliminated during normalization, while references to the translation of a type are retained in values of kind $\kity$. The key rules in the dynamics are: 
\begin{mathpar}
\small
\inferrule[s-ity-unquote-elim]{
    \sval{\sqity{\qity}}{\argEnv}
}{
    \sstep{\sqity{\qtuq{\sqity{\qity}}}}{\argEnv}{\sqity{\qity}}
}

\inferrule[s-ity-trans-val]{
    \sval{\st}{\argEnv}
}{
    \sval{\sqity{\srep{\st}}}{\argEnv}
}
\end{mathpar}

These are used in the schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g. we could also have used a list). We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}~\keyw{in}~\omega_\text{lprod}$  where $\stx{lprod/trans}:=$
\[\small
\begin{array}{l}
\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
\end{array}\]
Evaluating this translation schema with the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$. Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away

@@

Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL, which is intrinsically rather than bidirectionally typed). We use Plotkin's fixpoint operator for general recursion (cf. \cite{pfpl}), and define both lambdas and fixpoints only in analytic positions for simplicity.

\subsection{Generalized Introductory Operations}\label{sec:introop}
The translation of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.

@@

For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
\[\small\begin{array}{l}
    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\svar{str\_of\_Str}~\svar{tmidx}
}}}
\end{array}\]
Because regular strings are implemented as strings, this intro opcon definition is straightforward. It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$. In practice, the tycon provider would specify an error message here. 
Next, it checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure. Finally, a {translation} corresponding to the term index is generated via  helper function $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$.% and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.

>>>>

\AtBeginDocument{%
 \abovedisplayskip=5pt
 \abovedisplayshortskip=4pt
 \belowdisplayskip=4pt
 \belowdisplayshortskip=4pt
}

@@

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to use the same parametricity results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We conclude with  related and future work in Sec. \ref{prior-work}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent
These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only in the synthetic judgement.

@@

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation.

%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}. The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
%  \begin{figure*}[t!]
% \small
% \hspace{-5px}\input{example-table}
% %\\~\\
% %$\begin{array}{rcl}
% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
% %\end{array}$
% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
% \label{example}
% \end{figure*}

\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-EL}
\caption{Syntax of the external language (EL).}
\label{syntax-EL}\vspace{-5px}
\end{figure}

@@

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined. All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}. The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}\vspace{-3px}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}\vspace{-8px}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}\vspace{-8px}
\end{figure*}

@@

\paragraph{Examples}
Our first example of a user-defined tycon is $\tcvar{rstr}$. It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision. }

@@

\subsection{Static Type Case Analysis}
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-10px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-8px}
\end{figure}

Types in @$\lambda$ can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context \cite{conf/ppdp/LohH06}. Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. \vspace{-5px}
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}
The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (such types always take the default branch).

The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

@@

\inferrule[k-ity-trans]{
    \sofkX{\st}{\kty}
}{
    \sofkX{\sqity{\srep{\st}}}{\kity}
}
\end{mathpar}
The unquote form is eliminated during normalization, while references to type translations are retained in values:
\begin{mathpar}
\small
\inferrule[s-ity-unquote-elim]{
    \sval{\sqity{\qity}}{\argEnv}
}{
    \sstep{\sqity{\qtuq{\sqity{\qity}}}}{\argEnv}{\sqity{\qity}}
}

\inferrule[s-ity-trans-val]{
    \sval{\st}{\argEnv}
}{
    \sval{\sqity{\srep{\st}}}{\argEnv}
}
\end{mathpar}
These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g. we could also have used a list). We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}~\keyw{in}~\omega_\text{lprod}$  where $\stx{lprod/trans}:=$
\[\small
\begin{array}{l}
\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
\end{array}\]
Applying this translation schema to the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$. Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away

@@

Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL). We use Plotkin's fixpoint operator for general recursion (cf. \cite{pfpl}), and define both lambdas and fixpoints only analytically for simplicity.

\subsection{Generalized Introductory Operations}\label{sec:introop}
The translation of the \emph{generalized intro operation}, written $\small\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.

@@

For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
\[\small\begin{array}{l}
    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\svar{str\_of\_Str}~\svar{tmidx}
}}}
\end{array}\]
Because regular strings are implemented as strings, this intro opcon definition is straightforward. It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$. In practice, the tycon provider would specify an error message here. 
Next, it checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure. Finally, the \emph{translational internal string} corresponding to the static string provided as the term index is generated via the helper function $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$.% and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.
